Well-quasi-orderings on word languages

### 2024-10-17 src/introduction.tex
trying to update intro, in a bad way
new intro héhé

A \intro{well-quasi-ordered} set is a set $X$ equipped with a quasi-order
$\preceq$ such that every infinite sequence $\seqof{x_n}$ of elements taken in
$X$ contains a pair $x_i \preceq x_j$ with $i < j$. Well-quasi-orderings serve
as a core combinatorial tool powering many termination arguments, and was
successfully applied to the verification of infinite state transition systems
\cite{ABDU96,ABDU98}. One of the appealing properties of well-quasi-orderings
is that they are closed under many operations, such as taking products, finite
unions, and finite powersets \cite{SCSC12}. Perhaps more surprisingly, the
class of well-quasi-ordered sets is also stable under the operation of taking
finite words and finite trees labelled by elements of a well-quasi-ordered set
\cite{HIG52,KRU72}.

\AP Note that in the case of finite words and finite trees, the precise choice
of ordering is crucial to ensure that the resulting structure is
\kl{well-quasi-ordered}. The celebrated result of Higman states that the set of
finite words over an ordered alphabet $(X, \preceq)$ is \kl{well-quasi-ordered}
by the so-called \kl{subword embedding relation} \cite{HIG52}. Let us recall
that the \kl{subword relation} for words over $(X, \preceq)$ is defined as
follows: a word $u$ is a \intro{subword} of a word $v$, written $u
\intro*\higleq v$, if there exists an increasing function $f \colon \{1,
\ldots, |u|\} \to \{1, \ldots, |v|\}$ such that $u_i \preceq v_{f(i)}$ for all
$i \in \{1, \ldots, |u|\}$.

@@

\AP However, there are many other natural orderings on words that could be
considered in the context of \kl{well-quasi-orderings}. Let us restrict
ourselves to a simplified case where the alphabet is finite and equipped with
the equality relation. In this setting, the three alternatives we consider are
the \intro{prefix relation} ($u \intro*\prefleq v$ if there exists $w$ with $uw
= v$), the \intro{suffix relation} ($u \intro*\suffleq v$ if there exists $w$
such that $wu = v$), and the \intro{infix relation} ($u \intro*\infleq v$ if
there exists $w_1,w_2$ such that $w_1 u w_2 = v$). Note that these three
relations straightforwardly generalize to infinite quasi-ordered alphabets.
Unfortunately, it is easy to see that all of these constructions are not
well-quasi-ordered as soon as the alphabet contains two non-equivalent letters:
for instance, the infinite sequence $a^n b^n a^n$ is \kl{well-quasi-ordered} by
the \kl{subword relation} but by neither the \kl{prefix relation}, nor the
\kl{suffix relation}, nor the \kl{infix relation}.

\AP While this dooms \kl[well-quasi-ordered]{well-quasi-orderedness} of these relations in the general case, there may be
\emph{subsets} of $\Sigma^*$ which are \kl{well-quasi-ordered} by these
relations. As a simple example, take the case of finite sets of (finite) words
which are all \kl{well-quasi-ordered} regardless of the ordering considered.
This raises the question of characterizing exactly which subsets $L \subseteq \Sigma^*$
are \kl{well-quasi-ordered} with respect to the \kl{prefix relation}
(respectively, the \kl{suffix relation} or the \kl{infix relation}), and
providing suitable decision procedures.

\subparagraph{Contribution} In this paper, we focus on words over a finite
alphabet $\Sigma$, and characterize subsets $L \subseteq \Sigma^*$ that are
\kl{well-quasi-ordered} by the \kl{prefix relation}, the \kl{suffix relation},
and the \kl{infix relation}. Specifically, for the \kl{prefix relation} and the \kl{suffix relation}, we are able to show the following decomposition into \intro{chains}, that is, totally ordered ascending sequences.

@@

Note that \kl{chains} are the simplest possible
\kl{well-quasi-ordered} sets (they are totally ordered and well-founded) and it
is known that finite unions of \kl{well-quasi-ordered} sets are
\kl{well-quasi-ordered}. As a consequence, the above characterization states
that only the simplest possible \kl{well-quasi-ordered} sets are
\kl{well-quasi-ordered} by the \kl{prefix relation}. Let us furthermore
highlight that this characterization holds without any restriction on the
decidability of the language $L$ itself, but heavily relies on the assumption
that $\Sigma$ is finite.

\AP One way to measure the complexity of a \kl{well-quasi-ordered} set is to
consider its so-called \intro{ordinal invariants}. For instance, the
\intro{order type} of a \kl{well-quasi-ordered} set is the smallest ordinal
that cannot be embedded into it. For the \kl{prefix relation} and the \kl{infix
relation}, we conclude that the order type of a \kl{well-quasi-ordered} set is
at most $k \times \omega$, and its width is finite.\todo{check this + cite}.

\AP The straightforward generalization of the results for the \kl{prefix
relation} and the \kl{suffix relation} to the \kl{infix relation} is not
possible. For example, Kuske has shown that any countable partial order with
finite downward closures can be embedded into
it~\cite{DBLP:journals/ita/Kuske06}. In particular, this means that the
\kl{subword relation} can be embedded into the \kl{infix relation}. The former
being a \kl{well-quasi-ordered} with \kl{order type} $\omega^\omega$
\todo{cite}. However, not all is lost: we show that with some mild restrictions
we can recover much of the simple structure of the \kl[prefix relation]{prefix}
and \kl{suffix relation}:

{
\renewcommand{\labelenumi}{R\arabic{enumi}}
\begin{enumerate}
	\setcounter{enumi}{2}
    \item A \emph{bounded} language $L \subseteq \Sigma^*$ is \kl{well-quasi-ordered} by the \kl{infix relation} if and only if it is a finite union of languages $S_i \cdot P_i$, where each $S_i$ is \kl{well-quasi-ordered} by the \kl{suffix relation} and each $P_i$ is \kl{well-quasi-ordered} by the \kl{prefix relation}.
\end{enumerate}
}

We illustrate how this result can be used to study languages recognized by
finite automata, context-free grammars, petri nets, and other systems. This is
possible by leveraging the recently introduced notion of \kl{amalgamation
systems} \cite{ASZZ24}, which abstracts the notion of systems whose runs can be
suitably ``pumped'' and include common language classes such as regular
languages, context-free languages, and even languages recognized by petri nets.

{
\renewcommand{\labelenumi}{R\arabic{enumi}}
\begin{enumerate}
	\setcounter{enumi}{2}
	\item A language $L \subseteq \Sigma^*$ recognized by an \kl{amalgamation system} is \kl{well-quasi-ordered} by the \kl{infix relation} if and only if it is a finite union of languages $S_i \cdot P_i$, where each $S_i$ is \kl{well-quasi-ordered} by the \kl{suffix relation} and each $P_i$ is \kl{well-quasi-ordered} by the \kl{prefix relation}.
\end{enumerate}
}

Finally, there has been recent interest in deciding whether a given order is a
well-quasi-order. These questions have been investigated for example in the
context of the verification of counter
machines~\cite{DBLP:conf/fsttcs/FinkelG19} and in the context of
logic~\cite{DBLP:journals/pacmpl/BergstrasserGLZ24}. We conclude our results
with a contribution in this continuity.

\todo{verify this result :)}
{
\renewcommand{\labelenumi}{R\arabic{enumi}}
\begin{enumerate}
	\setcounter{enumi}{3}
	\item For languages of amalgamation systems, it is decidable whether they are \kl{well-quasi-ordered} by the \kl[prefix relation]{prefix}, \kl[suffix relation]{suffix}, or \kl{infix relation}.
\end{enumerate}
}

>>>>

\AP A \intro{well-quasi-ordered} set is a set $X$ equipped with a quasi-order
$\preceq$ such that every infinite sequence $\seqof{x_n}$ of elements taken in
$X$ contains a pair $x_i \preceq x_j$ with $i < j$. Well-quasi-orderings serve
as a core combinatorial tool powering many termination arguments, and was
successfully applied to the verification of infinite state transition
systems~\cite{ABDU96,ABDU98}. One of the appealing properties of
well-quasi-orderings is that they are closed under many operations, such as
taking products, finite unions, and finite powerset
constructions~\cite{SCSC12}. Perhaps more surprisingly, the class of
well-quasi-ordered sets is also stable under the operation of taking finite
words and finite trees labelled by elements of a well-quasi-ordered set
\cite{HIG52,KRU72}.

\AP Note that in the case of finite words and finite trees, the precise choice
of ordering is crucial to ensure that the resulting structure is
\kl{well-quasi-ordered}. The celebrated result of Higman states that the set of
finite words over an ordered alphabet $(X, \preceq)$ is \kl{well-quasi-ordered}
by the so-called \kl{subword embedding relation}~\cite{HIG52}. Let us recall
that the \kl{subword relation} for words over $(X, \preceq)$ is defined as
follows: a word $u$ is a \intro{subword} of a word $v$, written $u
\intro*\higleq v$, if there exists an increasing function $f \colon \{1,
\ldots, |u|\} \to \{1, \ldots, |v|\}$ such that $u_i \preceq v_{f(i)}$ for all
$i \in \{1, \ldots, |u|\}$.

@@

\AP However, there are many other natural orderings on words that could be
considered in the context of \kl{well-quasi-orderings}, even in the simplified
setting of a finite alphabet $\Sigma$ equipped with the equality relation. In
this setting, the three alternatives we consider are the \intro{prefix
relation} ($u \intro*\prefleq v$ if there exists $w$ with $uw = v$), the
\intro{suffix relation} ($u \intro*\suffleq v$ if there exists $w$ such that
$wu = v$), and the \intro{infix relation} ($u \intro*\infleq v$ if there exists
$w_1,w_2$ such that $w_1 u w_2 = v$). Note that these three relations
straightforwardly generalize to infinite quasi-ordered alphabets.
Unfortunately, it is easy to see that all of these constructions are not
well-quasi-ordered as soon as the alphabet contains two distinct letters:
for instance, the infinite sequence $a^n b^n a^n$ is \kl{well-quasi-ordered} by
the \kl{subword relation} but by neither the \kl{prefix relation}, nor the
\kl{suffix relation}, nor the \kl{infix relation}.

\AP While this dooms \kl[well-quasi-ordered]{well-quasi-orderedness} of these
relations in the general case, there may be \emph{subsets} of $\Sigma^*$ which
are \kl{well-quasi-ordered} by these relations. As a simple example, take the
case of finite sets of (finite) words which are all \kl{well-quasi-ordered}
regardless of the ordering considered. This raises the question of
characterizing exactly which subsets $L \subseteq \Sigma^*$ are
\kl{well-quasi-ordered} with respect to the \kl{prefix relation} (respectively,
the \kl{suffix relation} or the \kl{infix relation}), and together with
suitable decision procedures.

\AP Let us argue that this question is not only of theoretical interest.
Indeed, there has been recent breakthroughs in deciding whether a given order
is a \kl{well-quasi-order}, for instance in the context of the verification of
infinite state transition systems~\cite{DBLP:conf/fsttcs/FinkelG19} or in the
context of logic~\cite{DBLP:journals/pacmpl/BergstrasserGLZ24}. Our decision
procedures clearly contribute to this line of research in the context of words.
Furthermore, a previous work by Kuske shows that any
\emph{reasonable}\footnote{ This will be made precise in
\cref{infix-embedding:thm}. } partially ordered set $(X, \leq)$ can
be embedded into $\set{a,b}^*$ with the \kl{infix relation}~\cite[Lemma
5.1]{DBLP:journals/ita/Kuske06}. Phrased differently, one can encode a large
class of partially ordered sets as subsets of $\set{a,b}^*$. As a consequence,
the following decision problem provides a reasonable abstract framework for
deciding whether a given partially ordered set is \kl{well-quasi-ordered}:
given a language $L \subseteq \Sigma^*$, decide whether $L$ is
\kl{well-quasi-ordered} by the \kl{suffix relation}, or the \kl{infix
relation}.

\AP When considering an algorithm based on a \kl{well-quasi-ordering}, the
runtime of the algorithm is deeply related to the ``complexity'' of the
underlying quasi-order \todo{CITE???}. One way to measure this complexity is to
consider its so-called \intro{ordinal invariants}: for instance, the
\intro{maximal order type} (or m.o.t.), originally defined by De Jongh and
Parikh \cite{dejongh77}, is the order type of the maximal linearization of a
well-quasi-ordered set. In the case of a finite set, the m.o.t. is precisely
the size of the set. Better runtime bounds were obtained by considering two
other parameters\todo{cite sylvain width/ HDR?}: the \intro{ordinal height}
introduced by Schmidt \cite{schmidt81}, and the \intro{ordinal width} of Kříž
and Thomas \cite{kriz90b}. Therefore, when characterizing
\kl{well-quasi-ordered} languages, we will also be interested in computing
bounds on their \kl{ordinal invariants}. We refer to the
\cref{ordinal-invariants:subsec} for a more detailed introduction to these
parameters and ordinal computations in general. For the moment, let us just
remark that all languages $L \subseteq \Sigma^*$ have an \kl{ordinal height} of
at most $\omega$.

\subparagraph{Contribution} In this paper, we focus on words over a finite
alphabet $\Sigma$, and characterize subsets $L \subseteq \Sigma^*$ that are
\kl{well-quasi-ordered} by the \kl{prefix relation}, the \kl{suffix relation},
and the \kl{infix relation}. Furthermore, we devise decision algorithms
whenever the languages are given by reasonable computational models. Finally,
we provide relatively low bounds on the possible \kl{ordinal invariants} of
such languages.

Specifically, for the \kl{prefix relation} and the
\kl{suffix relation}, we are able to show the following decomposition into
\intro{chains}, that is, totally ordered ascending sequences.

@@

Note that \kl{chains} are the simplest possible \kl{well-quasi-ordered} sets
(they are totally ordered and well-founded) and it is known that finite unions
of \kl{well-quasi-ordered} sets are \kl{well-quasi-ordered}. As a consequence,
the above characterization states that only the simplest possible
\kl{well-quasi-ordered} sets are \kl{well-quasi-ordered} by the \kl{prefix
relation}. Let us furthermore highlight that this characterization holds
without any restriction on the decidability of the language $L$ itself, but
heavily relies on the assumption that $\Sigma$ is finite. This allows us to
derive the following bounds on the \kl{ordinal invariants} of such
\kl{well-quasi-ordered} languages, which can be interpreted in two dual ways:
first, these languages are extremely simple, which means that one could have
hoped for using directly another well-quasi-ordered set without resorting to
finite words; second, any time one encounters a \kl{well-quasi-ordered}
language, it is going to provide relatively efficient algorithms.

\renewcommand{\labelenumi}{R\arabic{enumi}}
\begin{enumerate}
    \setcounter{enumi}{2}
    \item All languages $L \subseteq \Sigma^*$ are 
        \kl{well-quasi-ordered} by the \kl{prefix relation} 
        (resp. \kl{suffix relation}) have an \kl{ordinal height} of at most $\omega$,
        a finite \kl{ordinal width}, and a \kl{maximal order type} strictly
        below $\omega^2$.
\end{enumerate}
}

\AP The straightforward generalization of the results for the \kl{prefix
relation} and the \kl{suffix relation} to the \kl{infix relation} is not
possible. Indeed, it follows from Kuske's result that $\Sigma^*$ equipped with
the \kl{subword relation} can be embedded into $\set{a,b}^*$ with the \kl{infix
relation}~\cite[Lemma 5.1]{DBLP:journals/ita/Kuske06}. This implies that there
are \kl{well-quasi-ordered} languages for the \kl{infix relation} that have
very large \kl{ordinal invariants}: for instance, the \kl{maximal order type}
of the \kl{subword relation} is $\omega^{\omega^{\card{\Sigma} - 1}}$, which
equals its \kl{ordinal width} \cite[Corollary 3.9, Theorem 4.21]{DZSCSC20}.
However, not all is lost if we restrict ourselves to \kl{bounded languages},
i.e., languages that are included in some $w_1^* \cdots w_k^*$ for some finite
choice of words $w_1, \ldots, w_k$. We are able to show the following:

{
\renewcommand{\labelenumi}{R\arabic{enumi}}
\begin{enumerate}
	\setcounter{enumi}{3}
    \item A \kl{bounded language} $L \subseteq \Sigma^*$ is \kl{well-quasi-ordered} by the \kl{infix relation} if and only if it is a finite union of languages $S_i \cdot P_i$, where each $S_i$ 
        is a \kl{chain} for the \kl{suffix relation},
        and where
        each $P_i$ is a \kl{chain} for the \kl{prefix relation}.
\end{enumerate}
}

This result directly translates into bounds on the possible \kl{ordinal invariants}
of such languages similarly as for the \kl{prefix relation}.
{
\renewcommand{\labelenumi}{R\arabic{enumi}}
\begin{enumerate}
    \setcounter{enumi}{4}
    \item All \kl{bounded languages}
            $L \subseteq \Sigma^*$ that are \kl{well-quasi-ordered} by the \kl{infix relation}
            have an \kl{ordinal height} of at most $\omega$,
            an \kl{ordinal width} strictly below $\omega^2$, and
            a \kl{maximal order type} strictly below $\omega^3$.
\end{enumerate}
}

\AP
Finally, we provide a decision procedure for checking whether a language is
\kl{well-quasi-ordered}, which takes as inputs languages recognized
by \kl{amalgamation systems} \cite{ASZZ24}. Such systems will be
formally introduced in \cref{amalgamation-systems:subsec}, but for the moment
let us just say that they include many classical computational models such as
finite automata, context-free grammars, and petri nets \cite{ASZZ24}. 
{
\renewcommand{\labelenumi}{R\arabic{enumi}}
\begin{enumerate}
    \setcounter{enumi}{5}
    \item There is a decision procedure for checking whether a language $L$ recognized by an \kl{amalgamation system} is \kl{well-quasi-ordered} by the \kl{prefix relation}, the \kl{suffix relation}, or the \kl{infix relation}. Furthermore, if it is \kl{well-quasi-ordered}, then $L$ is a \kl{bounded language}.
\end{enumerate}
}
Notice that the above result implies that the hypothesis of
\kl{bounded languages} on the theoretical side is not a restriction at all in
practice.
